A portable, annotation-based, visual stepper for Common Lisp

### 2020-03-06 paper/paper.tex
Update paper.tex

A program stepper lets users execute code, section by section, in
steps of their own choosing.  The stepper system automatically inserts
a number of hidden control points along the program's paths.  At each
point, the system may wait for user instructions before and/or after
executing the next section.  Steppers often fall within the category
of \emph{program execution monitoring}, along with profilers, tracers
and other debugging tools.  Stepping is the one of the most popular
forms of debugging since it allows users to study the evolution of a
program's state by direct inspection, as opposed to a combination of
conjecture and experimentation.  A stepper gives users full control
over the speed of the actual program and often even allows
manipulations of the program state, which frequently presents a
considerable debugging advantage over guessing the program's state or
monitoring its outputs.

Many programming systems, indeed all of the most popular programming
systems provide stepper tools as a part of a debugging tool-chain,
i.e. a set of programs often developed and distributed alongside
language compilers and interpreters.  In languages such as C, the
prevalence of a stepping system (such as the popular \texttt{gdb}
program) is so strong that it becomes synonymous with the term
``debugger'', which often leads to users migrating to the Lisp family
of languages being surprised to find debugging systems that aren't
about stepping at all.

@@

Common Lisp programmers are so detached from the practice of stepping
that some will simply declare that's proof that a stepper is useless.
There is a hint of truth to this, since efforts have been directed to
other types of debugging facilities such as powerful interactive
program restarts and function traces.  At any rate, it seems
undeniable that, stepper or no stepper, Common Lisp programs will be
debugged.

@@

Some Lisp implementations can and do implement stepping to various
levels of ability.  SBCL combines its implementation of
\texttt{CL:STEP} with the restarts system to provide a text-based
stepper, while the LispWorks\cite{lispworks-stepper} and Allegro CL
systems have sophisticated graphical stepping dialogues with views to
the source code and program state .  Unfortunately, these systems are
unavailable to whomever wishes to try portable programs across
different Common Lisp implementations.  Even if one switches between
two implementations that do have steppers, the difference in
capability and user interface is often enough to discourage either
system's use.

Interface inconsistency between implementations is not an exclusive
problem of the stepper feature: other debugging tools such as the
inspector, the debugger or the REPL suffer from it.  Thus, many Common
Lisp programmers will use a generic text editor such as GNU
Emacs\cite{emacs} together with the SLIME\cite{slime} extension.  This
combination forms a fully capable, implementation-agnostic Common Lisp
IDE that suppresses the problems described above and provides a
consistent user interface to many debugging features.

Regrettably, even though Emacs provides visual stepper interfaces for
many programming languages, SLIME doesn't provide a portable stepper
interface.  We believe this to be due to the fact that the technical
challenges to be surmounted are greater than for other debugging
tools.  Among other problems, the portability mandate of SLIME implies
it is ultimately only allowed to invoke functionality present in all
the Common Lisp systems it connects to.  This means that even if all
implementations where to implement some form of \texttt{CL:STEP}, that
alone wouldn't be powerful enough to, say communicate source location
information to and from Emacs, annotating the program source.

@@

In fact, the SLY Common Lisp IDE\cite{sly}, a derivative program of
SLIME, has redesigned some of its predecessor's underpinnings to make
this even easier.  We shall describe how ``stickers'', a feature that
SLY has recently acquired, lets users manually instrument selected
Common Lisp forms whose results they are interested in.  Stickers are
already a ``poor man's stepper'', in the sense that they have the
fundamental semantics of stepping but still encumber the user with
work than could be performed automatically.  To fill this gap, we
shall explore methods of combining stickers with automatic code
analysis.  We shall then be able to present an innovative stepper tool
for the Common Lisp language based on the SLY extension to the Emacs
editor, hereafter designated the \emph{SLY/Emacs stepper}.

@@

The rest of this essay is structured as follows: we start by reviewing
related work, so as to establish better context for the difficulties
encountered and decisions made.  We'll then describe the three main
components of the proposed SLY/Emacs stepper: (1) the interface
between SLY/Emacs stickers and Common Lisp; (2) a source-tracking form
reader; and (3) a manner of macroexpanding code that maintains the
memory of source-tracked forms.  We shall finish with a discussion of
the stepper tool's interface, its limitations and future work.

@@

As can be seen, every time the expression above is executed, expansion
of the wrapper macro leads to the functions \texttt{before-sticker}
and \texttt{after-sticker} being called with the appropriate
\texttt{\%sticker} object.  Depending on the user's preference, these
functions may decide to stop execution of the program (by way of
\texttt{invoke-debugger}), or to simply record the fact that the
sticker was traversed.  The list of recordings can later be retrieved
and replayed later inside SLY/Emacs.

In our simple example, the benevolent user placed stickers on two
expressions that are indeed execute, i.e. they are exist in places of
evaluation as defined by the syntax of the \texttt{(let ...)}  special
form and the \texttt{(+ ...)} function form.  If a sticker had instead
been placed on the expressions \texttt{((baz 42))} or \texttt{+} --
two examples of non-evaluated forms -- that would have created a
difficulty, since the code would become syntactically invalid and fail
compilation.  In a worse situation, the code would still be
syntactically valid but semantically absurd.

SLY/Emacs's sticker system doesn't have a way to reject these
individual stickers, so it proceeds heuristically: it rejects the
compilation of the whole top-level form if it determines that arming
stickers results in an \emph{increase} to the number of compilation
warnings.  In that case, the original form is compiled but the
stickers \emph{fail to arm}.  This method work for a vast majority of
cases, but it doesn't seem impossible to construct a pathological case
where the principle of functional equivalence stated above is
violated.

@@

At its simplest definition, a source-tracking form reader is a
variation of the Common Lisp \texttt{CL:READ} function that invokes a
hook every time a sub-form is read, and proceeds to pass to this hook
a measure of the character distance traveled so far to read it.  This
enables problems that need both the usual results of \texttt{CL:READ}
and a table of form-to-source-code pairings.

@@

\begin{enumerate}
\item \texttt{Eclector}\cite{eclector}, a self-described ``portable
  Common Lisp reader that is highly customizable and can return
  concrete syntax trees'', is a full realization of a Common Lisp code
  reader that doesn't rely on the implementation's.  Its
  distinguishing characteristic is the fact that ``concrete syntax
  trees'' are not \texttt{CONS} cells, rather CLOS objects that mimic
  the properties of such cells while also keeping ``concrete''
  source file information;

@@

We reasoned that our stepper tool already expects a compliant Common
Lisp implementation, therefore it may, by definition, also expect a
compliant \texttt{CL:READ}.  So we set out to design a portable
source-tracking reader that completely reuses the implementation's
reader instead of replacing its implementation entirely.  To achieve
this, we settled on an arrangement of two separate techniques:

@@

\item A \emph{substitute read-table}, achieved by rebinding the
  variable \texttt{*READTABLE*}.  This table \emph{shadows} each of
  the entries of the current read-table (using
  \texttt{GET/SET-MACRO-CHARACTER}) with a function that allows full
  control of each characters influence in the returned symbolic
  expression.  By setting up this function in a particular manner, the
  resulting table can be made functionally indistinguishable from the
  original one, while still having the ability to invoke a user's
  hook to record source positions.
\end{itemize}

@@

This mechanism provides a simple, reliable\footnote{This was tested in
  SBCL, Allegro CL, CCL and ECL.  There are differences to the way
  that some implementations will construct the standard read-table
  (for example, SBCL and Allegro CL represent constituent characters
  differently) and \texttt{SUBSTITUTION-TABLE} has special provisions
  for that.  The same technique should in theory work with
  non-standard read-tables, but this has not been tested.} and
portable\footnote{As noted, except for the use of gray streams.}  way
to read source-code.  We can demonstrate its use on the simple
\texttt{(let ...)} form already presented above:

@@

To reach a state where nothing but special and function forms exist, a
macroexpander must remove macro by expanding them.  However, in doing
so, our program \emph{must also remember} whence each macro's
expansion came, specifically the source position of the form in its
pre-expansion state.  This what we mean by \emph{mnesic
  macroexpansion}, the opposite of \emph{amnesic macroexpansion}

@@

By this point, the system may finally come to the realization that
only the forms \texttt{42}, \texttt{baz}, \texttt{(plusp baz)},
\texttt{(foo)}, \texttt{(bar)}, \texttt{(if ...)} and \texttt{(let
  ...)} are in positions of evaluation.  Regrettably, it may now have
lost track as to \emph{where} each form lives in the source.

To recover this information, it is not enough to naively consult the a
hash-table produced in \ref{st-reader}, since some forms didn't
exist in our original source-tracked version, and even if they did,
the macroexpanding facilities are not generally obliged to return the
same \texttt{CONS} objects for the forms, regardless of whether they
expand them or not.

@@

To overcome this obstacle we need a different approach.  Instead of
trying to recover from a fully macroexpanded tree, we must hook into
macroexpansion as soon as it happens.  This shall allow us to lose as
little source-tracking information as possible.  Thus we conclude that
a programmable, portable code-walker is necessary.  Such a system
shall let us execute a hook at each macroexpansion step shortly before
and shortly after each expansion.

After surveying open-source alternatives for code-walkers, we settle
on a program called \texttt{AGNOSTIC-LIZARD}\cite{agnostic-lizard},
which fits exactly these requirements\footnote{Furthermore,
  \texttt{AGNOSTIC-LIZARD} contains useful provisions to shield user
  code from certain nonconformities in the macroexpansions of certain
  built-in macros, such as \texttt{DEFUN}.}.  Its main primitive,
\texttt{AGNOSTIC-LIZARD:WALK-FORM}, produces the desired full
macroexpansion, and can also be given a set of callback functions as
hooks.

Here's the snippet that illustrates our use of this walker.

@@

As we are nearing the end of our journey and can now start putting all
the pieces together. The following snippet is the final form of our
Common Lisp stepper function, whose results can be handed to SLY/Emacs
for instrumentation through stickers as described in section
\ref{stickers}.

@@

We have released the result of our work on the GitHub
platform\footnote{See
  \texttt{git@github.com:joaotavora/sly-slepper.git}.}.

@@

We shall see as we discuss its limitations that the resulting
SLY/Emacs stepper tool is still in its infancy.  It can nevertheless
be said to work reasonably well for a majority of normal
circumstances, succeeding in instrumenting forms effectively and
efficiently, while providing satisfactory methods of navigation among
stickers.

@@

In this example, we note that the atom \texttt{X} has two different
\emph{manifestations} in the encompassing form.  Naturally one wants
only the latter to be annotated, and not the first.  However, this is
hard to determine reliably since both are represented by the very same
object.  This is in stark contrast to compound forms represented by
different \texttt{CONS} cells.

We can enhance the form/position pairings table used above to record
the fact that there is more than one manifestation of an atom, but
that's not enough to know in \texttt{MNESIC-MACROEXPAND-ALL} which of
those is a in a position of execution.  This is because the
\texttt{AGNOSTIC-LIZARD} macroexpander will only traverse sub-forms of
forms actually returned by a macro's expansion.  In this example, our
hook is only called on the \texttt{(lambda (x) x)} and \texttt{x}
forms, \emph{not} on the \texttt{(x)} form.  The latter form is merely
an argument to the macro itself where we have no power of
intervention, and thus there is no easy way to invalidate the first
manifestation.

@@

These features don't seem hard to realize.  E.g. to enable more
sophisticated navigation behavior behavior the \texttt{RECORD} macro
could use a special variable to be made aware of recursive invocations
to itself.  Thus we could keep track of stickers being executed
\emph{inside} each other, i.e. within the dynamic scope of a
previously active sticker annotation.

@@

Common Lisp users are concerned about features that facilitate
day-to-day development.  \texttt{TRACE}, REPLs, \texttt{PRINT}
statements, the interactive debuggers, profilers and stickers are all
ways to solve certain debugging problems: some are more suitable to
some situations than others.  Therefore, it's important to stress that
program stepping is just another tool in the toolbox, and not way a
panacea.

By the same token, if one does implement such a tool, it should be
done in a manner that is of actual, practical use.  This was the
thinking behind the SLY/Emacs stepper.  We think it especially
fortuitous that annotation-based approach and the manner of
source-code correlation in SLY/Emacs's stickers feature doesn't make
use of direct references to source files or file positions.  As was
shown in \ref{stickers}, we merely keep an enumerated list of stickers
identifiers synchronized between Common Lisp and SLY/Emacs, a
mechanism that is simple but effective.  It adequately resists some
modifications to the source of the instrumented form or its
whereabouts, such as moving it around in the source file, or even
adding white-space and comments inside it.  This detail is crucial in
making stickers and stepping actually usable for day-to-day
programming, since the user isn't dragged away from his editor or
forced to a special confinement while stepping.  We note that these
advantages of annotation-based steppers were already hinted at back in
1989 by the proponents of evaluation-based steppers\cite[l.41,
l.42]{evaluation-based}.

@@

We also think it fortuitous that stepping is implemented as an
increment to the existing stickers functionality.  Like in
\cite{annotation-based} we consider it an advantage of the
annotation-based stepping systems that users are allowed to instrument
only the definitions they are interested in.  Indeed, it is often the
case that stepper systems become tedious to operate because they step
on \emph{too much}.  Our system has an additional selectivity
advantage: since the annotations are stickers, users can manually
adjust the automatically placed stickers, removing the ones they are
not interested in, or adding others.

@@

We would like to thank Luís Oliveira for early comments and advice
when discussing the stepper feature, and Andrew Lawson for the
encouragement to produce this work.

>>>>

A program stepper lets users execute code, section by section, in
steps of their own choosing.  A number of hidden control points are
inserted along the program's execution paths.  At each point, the
system may interrupt the program and wait for user instructions before
and/or after executing the next section.  Steppers often fall within
the category of \emph{program execution monitors}, along with
profilers, tracers and other debugging tools.  Stepping is the one of
the most popular forms of debugging since it allows users to study the
evolution of a program's state by direct inspection, as opposed to a
combination of conjecture and experimentation.  A stepper gives users
full control over the speed of the actual program and often even
allows manipulations of the program's state.  This frequently presents
a debugging advantage over guessing the program's state or monitoring
its outputs.

Many programming systems, indeed all of the most popular programming
systems provide stepper tools as a part of a debugging tool-chain,
i.e. a set of programs often developed and distributed alongside
language compilers and interpreters.  In languages such as C, the
prevalence of a stepping system (such as the popular \texttt{gdb}
program) is so strong that it becomes synonymous with the term
``debugger''. This sometimes leads to users migrating to the Lisp
family of languages being surprised to find debugging systems that
aren't about stepping at all.

@@

Common Lisp programmers are so detached from the practice of stepping
that some will simply declare that's proof that they don't need one at
all.  There is a hint of truth to this, as Common Lisp systems have
traditionally directed efforts to other types of debugging facilities
such as powerful interactive program restarts and function traces.  At
any rate, it seems undeniable that, stepper or no stepper, Common Lisp
programs will be debugged.

@@

Some Lisp implementations can and do implement stepping to various
levels of ability.  SBCL combines its implementation of
\texttt{CL:STEP} with the restarts system to provide a text-based
stepper, while the LispWorks\cite{lispworks-stepper} and Allegro CL
systems have sophisticated graphical stepping dialogues with views to
the source code and program state.  Unfortunately, these systems are
unavailable to whomever wishes to try portable programs across
different Common Lisp implementations.  Even if one switches between
two implementations that do have steppers, the difference in
capability and user interface is often enough to discourage either
system's use.

The problem of interface inconsistency across implementations is not
exclusive to the stepper feature: other debugging tools such as the
inspector, the debugger or the REPL suffer from it.  Thus, many Common
Lisp programmers will use a generic text editor such as GNU
Emacs\cite{emacs} together with the SLIME\cite{slime} extension.  This
combination forms a capable, implementation-agnostic Common Lisp IDE
that suppresses the problems described above and provides a consistent
user interface to many debugging features.

Regrettably, even though Emacs provides visual stepper interfaces for
many programming languages, SLIME doesn't provide a portable stepper
interface.  We believe this to be due to the fact that the technical
challenges to be surmounted are greater than for other debugging
tools.  Among other problems, the portability mandate of SLIME implies
it is ultimately only allowed to invoke functionality present in all
the Common Lisp systems it connects to.  This means that even if all
implementations where to implement some form of \texttt{CL:STEP}, that
alone wouldn't be powerful enough to, say, communicate source location
information to and from Emacs, annotating the program source.

@@

In fact, the SLY Common Lisp IDE\cite{sly}, a derivative program of
SLIME, has redesigned some of its predecessor's underpinnings to make
this even easier.  We shall describe how ``stickers'', a feature that
SLY has recently acquired, lets users manually instrument selected
Common Lisp forms whose results they are interested in.  Stickers are
already a ``poor man's stepper'', in the sense that they have some
fundamental semantics of stepping but still encumber the user with
work than could be performed automatically.  To fill this gap, we
shall explore methods of combining stickers with automatic code
analysis.  We shall then be able to present an innovative stepper tool
for the Common Lisp language based on the SLY extension to the Emacs
editor, hereafter designated the \emph{SLY/Emacs stepper}.

@@

The rest of this essay is structured as follows: we start by reviewing
related work, to establish better context for the difficulties
encountered and decisions made.  We'll then describe the three main
components of the proposed SLY/Emacs stepper: (1) the interface
between SLY/Emacs stickers and Common Lisp; (2) a source-tracking form
reader; and (3) a manner of macroexpanding code that maintains the
memory of source-tracked forms.  We shall finish with a discussion of
the stepper tool's interface, its limitations and future work.

@@

As can be seen, every time the expression above is executed, expansion
of the wrapper macro causes the functions \texttt{before-sticker} and
\texttt{after-sticker} to be called with the appropriate
\texttt{\%sticker} object.  Depending on the user's preference, these
functions may decide to stop execution of the program (by way of
\texttt{invoke-debugger}), or to simply record the fact that the
sticker was traversed.  The list of recordings can later be retrieved
and replayed later inside SLY/Emacs.

In our simple example, the benevolent user placed stickers on two
expressions that are indeed executed, i.e. they exist in places of
evaluation as defined by the syntax of the \texttt{(let ...)}  special
form and the \texttt{(+ ...)} function form.  If a sticker had instead
been placed on the expressions \texttt{((baz 42))} or \texttt{+} --
two examples of non-evaluated forms -- that would have created a
difficulty, since the code would become syntactically invalid and fail
compilation.  In a worse situation, the code would still be
syntactically valid but semantically absurd.

SLY/Emacs's sticker system doesn't have a way to reject these
individual stickers, so it proceeds heuristically: it rejects the
compilation of the whole top-level form if it determines that arming
stickers results in an \emph{increase} to the number of compilation
warnings.  In that case, the original form is compiled but the
stickers \emph{fail to arm}.  This strategy works for a vast majority
of cases, but it doesn't seem impossible to construct a pathological
case where the principle of functional equivalence stated above is
violated.

@@

At its simplest definition, a source-tracking form reader is a
variation of the Common Lisp \texttt{CL:READ} function that invokes a
hook every time a sub-form is read, and proceeds to pass to this hook
a measure of the character distance traveled so far to read it.  This
enables programs that need both the usual results of \texttt{CL:READ}
and a table of form-to-source-code pairings.

@@

\begin{enumerate}
\item \texttt{Eclector}\cite{eclector}, a self-described ``portable
  Common Lisp reader that is highly customizable and can return
  concrete syntax trees'', is a full realization of a Common Lisp code
  reader that doesn't rely on the implementation's.  Its
  distinguishing characteristic are ``concrete syntax trees'' that
  aren't represented by \texttt{CONS} cells, rather by CLOS objects
  that mimic the properties of such cells while also keeping
  ``concrete'' source file information;

@@

We reasoned that our stepper tool already expects a compliant Common
Lisp implementation.  Therefore it may, by definition, also expect a
compliant \texttt{CL:READ}.  So we set out to design a portable
source-tracking reader that completely reuses the implementation's
reader instead of replacing its implementation entirely.  To achieve
this, we settled on an arrangement of two separate techniques:

@@

\item A \emph{substitute read-table}, achieved by rebinding the
  variable \texttt{*READTABLE*}.  This table \emph{shadows} each of
  the entries of the current read-table (using
  \texttt{GET/SET-MACRO-CHARACTER}) with a function that allows full
  control of each characters influence in the returned symbolic
  expression.  By setting up this function in a particular manner, the
  resulting table can be made functionally indistinguishable from the
  original one, while retaining the ability to invoke a user's hook to
  record source positions.
\end{itemize}

@@

This mechanism provides a simple, reliable\footnote{This was tested in
  SBCL, Allegro CL, CCL and ECL.  There are differences to the way
  that some implementations will construct the standard read-table
  (for example, SBCL and Allegro CL represent ``constituent''
  characters differently) and \texttt{SUBSTITUTION-TABLE} has special
  provisions for that.  The same technique should in theory work with
  non-standard read-tables, but this has not been tested.} and
portable\footnote{As noted, except for the use of gray streams.}  way
to read source-code.  We can demonstrate its use on the simple
\texttt{(let ...)} form already presented above:

@@

To reach a state where nothing but special and function forms exist, a
macroexpander must remove macros by expanding them.  However, in doing
so, our program \emph{must also remember} whence each macro's
expansion came, specifically the source position of the form in its
pre-expansion state.  This what we mean by \emph{mnesic
  macroexpansion}, the opposite of \emph{amnesic macroexpansion}

@@

By this point, the system may finally come to the realization that
only the forms \texttt{42}, \texttt{baz}, \texttt{(plusp baz)},
\texttt{(foo)}, \texttt{(bar)}, \texttt{(if ...)} and \texttt{(let
  ...)} are in positions of evaluation.  Regrettably, it may now have
lost track of \emph{where} each form lives in the source.

To recover this information, it is not enough to naively consult the
hash-table produced in \ref{st-reader}, since some forms didn't exist
in our original source-tracked version.  Even if they did, the
macroexpanding facilities are not generally obliged to return the same
\texttt{CONS} objects for the forms, regardless of whether they expand
them or not.

@@

To overcome this obstacle we need a different approach.  Instead of
trying to recover from a fully macroexpanded tree, we must hook into
macroexpansion as soon as it happens.  This shall allow us to lose as
little source-tracking information as possible.  Thus, we conclude
that a programmable, portable code-walker is necessary.  Such a system
shall let us execute a hook at each macroexpansion step shortly before
and shortly after each expansion.

After surveying open-source alternatives for code-walkers, we settled
on a program called \texttt{AGNOSTIC-LIZARD}\cite{agnostic-lizard},
which fits exactly these requirements\footnote{Furthermore,
  \texttt{AGNOSTIC-LIZARD} contains useful provisions to shield user
  code from certain nonconformities in the macroexpansions of certain
  built-in macros, such as \texttt{DEFUN}.}.  Its main primitive,
\texttt{AGNOSTIC-LIZARD:WALK-FORM}, produces the desired full
macroexpansion, and can also be given a set of callback functions as
hooks.

Here's the snippet that illustrates our use of this walker:

@@

As we are nearing the end of our journey, we can now start putting all
the pieces together. The following snippet is the final form of our
Common Lisp function. Its results can be handed to SLY/Emacs for
instrumentation through stickers as described in section
\ref{stickers}.  After compilation, the instrumented code is now
steppable.

@@

We have released the result of our work on the GitHub
platform\footnote{See
  \texttt{https://github.com/joaotavora/sly-slepper.git}.}.

@@

We shall see, as we discuss its limitations, that the resulting
SLY/Emacs stepper tool is still in its infancy.  It can nevertheless
be said to work reasonably well for a majority of normal
circumstances, succeeding in instrumenting forms effectively and
efficiently, while providing satisfactory methods of navigation among
stickers.

@@

In this example, we note that the atom \texttt{X} has two different
\emph{manifestations} in the encompassing form.  Naturally one wants
only the latter to be annotated, and not the first.  However, that is
hard to determine reliably since both are represented by the very same
object.  This is in stark contrast to compound forms represented by
different \texttt{CONS} cells.

We can enhance the form/position pairings table used above to record
the fact that there is more than one manifestation of an atom, but
that's not enough to know in \texttt{MNESIC-MACROEXPAND-ALL} which of
those is a in a position of execution.  The reason is that the
\texttt{AGNOSTIC-LIZARD} macroexpander will only traverse sub-forms of
forms actually returned by a macro's expansion.  In this example, our
hook is only called on the \texttt{(lambda (x) x)} and \texttt{x}
forms, \emph{not} on the \texttt{(x)} form.  The latter form is merely
an argument to the macro itself where we have no power of
intervention, and thus there is no easy way to invalidate the first
manifestation.

@@

These features don't seem hard to realize.  E.g. to enable more
sophisticated navigation behavior behavior the \texttt{RECORD} macro
could use a special variable to be made aware of recursive invocations
to itself or to the currently executing stack frame.  Thus we could
keep track of stickers being executed \emph{inside} each other,
i.e. within the dynamic scope of a previously active sticker
annotation.

@@

Common Lisp users are concerned about features that facilitate
day-to-day development.  \texttt{TRACE}, REPLs, \texttt{PRINT}
statements, the interactive debuggers, profilers and stickers are all
ways to solve certain debugging problems: some are more suitable to
some situations than others.  Therefore, it's important to note that
program stepping is just another tool in the toolbox, not a
panacea.

By the same token, if one does implement such a tool, it should be
done in a manner that is of actual, practical use.  This was the
thinking behind the SLY/Emacs stepper.  We think it especially
fortunate that annotation-based approach and the manner of source-code
correlation in SLY/Emacs's stickers feature doesn't make use of direct
references to source files or file positions.  As was shown in
\ref{stickers}, we merely keep an enumerated list of stickers
identifiers synchronized between Common Lisp and SLY/Emacs, a
mechanism that is simple but effective.  It adequately resists some
modifications to the source of the instrumented form or its
whereabouts, such as moving it around in the source file, or even
adding white-space and comments inside it.  This detail is crucial in
making stickers and stepping actually usable for day-to-day
programming, since the user isn't dragged away from his editor or
forced to a special confinement while stepping.  We note that these
advantages of annotation-based steppers were already hinted at back in
1989 by the proponents of evaluation-based steppers\cite[l.41,
l.42]{evaluation-based}.

@@

We also think it fortunate that stepping is implemented as an
increment to the existing stickers functionality.  As in
\cite{annotation-based}, we consider it an advantage of the
annotation-based stepping systems that users are allowed to instrument
only the definitions they are interested in.  Indeed, it is often the
case that stepper systems become tedious to operate because they step
on \emph{too much}.  Our system has an additional selectivity
advantage: since the annotations are stickers, users can manually
adjust the automatically placed stickers, removing the ones they are
not interested in, or adding others.

@@

We would like to thank Luís Oliveira and Michael Kirkland for early
comments and advice when discussing the stepper feature, as well as
Andrew Lawson for the encouragement to produce this work.
